\chapter{Bytes objects}
\label{byteschapter}

Many applications must deal with blocks of binary data by accessing
them in various ways---extracting signed or unsigned numbers of
various sizes.  Therefore, the \deflibrary{r6rs bytes} library
provides a single type for
blocks of binary data with multiple ways to access that data. It deals
only with integers in various sizes with specified endianness, because
these are the most frequent applications.

Bytes objects\mainindex{bytes objects} are objects of a disjoint
type. Conceptually, a bytes object represents a sequence of 8-bit
bytes.  The description of bytes objects uses the term \defining{byte}
for an exact integer in the interval $\{-128, \ldots, 127\}$ and the
term \defining{octet} for an exact integer in the interval $\{0,
\ldots, 255\}$.  A byte corresponds to its two's complement
representation as an octet.

The length of a bytes object is the number of bytes it contains. This
number is fixed. A valid index into a bytes object is an exact,
non-negative integer. The first byte of a bytes object has index 0;
the last byte has an index one less than the length of the bytes
object.

Generally, the access procedures come in different flavors according
to the size of the represented integer, and the endianness of the
representation.  The procedures also distinguish signed and unsigned
representations.
The signed representations all use two's complement.

Like list and vector literals, literals representing bytes objects
must be quoted:
%
\begin{scheme}
'\#vu8(12 23 123) \ev \#vu8(12 23 123)%
\end{scheme}


\begin{entry}{%
\proto{endianness}{ {\cf big}}{\exprtype}
\rproto{endianness}{ {\cf little}}{\exprtype}}
   
{\cf (endianness big)} and {\cf (endianness little)} evaluate to the
symbols {\cf big} and {\cf little}, respectively. These symbols
represent an endianness, and whenever one of the procedures operating
on bytes objects accepts an endianness as an argument, that argument
shall be one of these symbols. It is a syntax violation for the operand to
{\cf endianness} to be
anything other than {\cf big} or {\cf little}.
\end{entry}

\begin{entry}{%
\proto{native-endianness}{}{procedure}}

Returns the implementation's preferred endianness
(usually that of the underlying machine architecture),
either {\cf big} or {\cf little}.
\end{entry}   

\begin{entry}{%
\proto{bytes?}{ obj}{procedure}}
   
Returns \schtrue{} if \var{obj} is a bytes object,
otherwise returns \schfalse{}.
\end{entry}

\begin{entry}{%
\proto{make-bytes}{ k}{procedure}
\rproto{make-bytes}{ k fill}{procedure}}
   
Returns a newly allocated bytes object of \var{k} bytes.
   
If the \var{fill} argument is missing, the initial contents of the
returned bytes object are unspecified.
   
If the \var{fill} argument is present, it shall be an exact integer in
the interval $\{-128, \ldots 255\}$ that specifies the initial value
for the bytes of the bytes object: If \var{fill} is positive, it is
interpreted as an octet; if it is negative, it is interpreted as a byte.
\end{entry}   

\begin{entry}{%
\proto{bytes-length}{ bytes}{procedure}}
   
Returns, as an exact integer, the number of bytes in \var{bytes}.
\end{entry}

\begin{entry}{%
\proto{bytes-u8-ref}{ bytes k}{procedure}
\proto{bytes-s8-ref}{ bytes k}{procedure}}
   
\domain{\var{K} shall be a valid index of bytes.}
   
The {\cf bytes-u8-ref} procedure returns the byte at index \var{k} of \var{bytes},
as an octet.
   
The {\cf bytes-s8-ref} procedure returns the byte at index \var{k} of \var{bytes},
as a (signed) byte.

\begin{scheme}
(let ((b1 (make-bytes 16 -127))
      (b2 (make-bytes 16 255)))
  (list
    (bytes-s8-ref b1 0)
    (bytes-u8-ref b1 0)
    (bytes-s8-ref b2 0)
    (bytes-u8-ref b2 0))) \ev (-127 129 -1 255)
\end{scheme}
\end{entry}   

\begin{entry}{%
\proto{bytes-u8-set!}{ bytes k octet}{procedure}
\proto{bytes-s8-set!}{ bytes k byte}{procedure}}
   
\domain{\var{K} shall be a valid index of \var{bytes}.}
   
The {\cf bytes-u8-set!} procedure stores \var{octet} in element \var{k} of
\var{bytes}.
   
The {\cf bytes-s8-set!} procedure stores the two's complement representation of
\var{byte} in element \var{k} of \var{bytes}.
   
Both procedures return \unspecifiedreturn.

\begin{scheme}
(let ((b (make-bytes 16 -127)))

  (bytes-s8-set! b 0 -126)
  (bytes-u8-set! b 1 246)

  (list
    (bytes-s8-ref b 0)
    (bytes-u8-ref b 0)
    (bytes-s8-ref b 1)
    (bytes-u8-ref b 1))) \ev (-126 130 -10 246)
\end{scheme}
\end{entry}

\begin{entry}{%
\proto{bytes-uint-ref}{ bytes k endianness size}{procedure}
\proto{bytes-sint-ref}{ bytes k endianness size}{procedure}
\proto{bytes-uint-set!}{ bytes k n endianness size}{procedure}
\proto{bytes-sint-set!}{ bytes k n endianness size}{procedure}}
   
\domain{\var{Size} shall be a positive exact integer. $\{\var{k}, \ldots,
  \var{k} + \var{size} - 1\}$ shall be valid indices of \var{bytes}.}
   
{\cf bytes-uint-ref} retrieves the exact integer corresponding to the
unsigned representation of size \var{size} and specified by \var{endianness}
at indices $\{\var{k}, \ldots, \var{k} + \var{size} - 1\}$.
   
{\cf bytes-sint-ref} retrieves the exact integer corresponding to the two's
complement representation of size \var{size} and specified by \var{endianness} at
indices $\{\var{k}, \ldots, \var{k} + \var{size} - 1\}$.
   
\domain{For {\cf bytes-uint-set!}, \var{n} shall be an exact integer in the
  set $\{0, \ldots, 256^{\var{size}}-1\}$.}

{\cf bytes-uint-set!} stores the unsigned representation of size \var{size}
and specified by \var{endianness} into \var{bytes} at indices
$\{\var{k}, \ldots, \var{k} + \var{size} - 1\}$.
   
\domain{For {\cf bytes-sint-set!}, \var{n} shall be an exact integer in
  the interval $\{-256^{\var{size}}/2, \ldots,
  256^{\var{size}}/2-1\}$.}
{\cf bytes-sint-set!} stores the two's complement
representation of size \var{size} and specified by \var{endianness}
into \var{bytes} at indices $\{\var{k}, \ldots, \var{k} + \var{size} - 1\}$.
   
The \ldots{\cf -set!} procedures return \unspecifiedreturn.

\begin{scheme}
(define b (make-bytes 16 -127))

(bytes-uint-set! b 0 (- (expt 2 128) 3)
                 (endianness little) 16)

(bytes-uint-ref b 0 (endianness little) 16)\lev
    \#xfffffffffffffffffffffffffffffffd

(bytes-sint-ref b 0 (endianness little) 16)\lev -3

(bytes->u8-list b)\lev (253 255 255 255 255 255 255 255
               255 255 255 255 255 255 255 255)

(bytes-uint-set! b 0 (- (expt 2 128) 3)
                 (endianness big) 16)

(bytes-uint-ref b 0 (endianness big) 16) \lev
    \#xfffffffffffffffffffffffffffffffd

(bytes-sint-ref b 0 (endianness big) 16) \lev -3

(bytes->u8-list b) \lev (255 255 255 255 255 255 255 255
               255 255 255 255 255 255 255 253))
\end{scheme}
\end{entry}

\begin{entry}{%
\proto{bytes-u16-ref}{ bytes k endianness}{procedure}
\proto{bytes-s16-ref}{ bytes k endianness}{procedure}
\proto{bytes-u16-native-ref}{ bytes k}{procedure}
\proto{bytes-s16-native-ref}{ bytes k}{procedure}
\proto{bytes-u16-set!}{ bytes k n endianness}{procedure}
\proto{bytes-s16-set!}{ bytes k n endianness}{procedure}
\proto{bytes-u16-native-set!}{ bytes k n}{procedure}
\proto{bytes-s16-native-set!}{ bytes k n}{procedure}}
   
\domain{\var{K} shall be a valid index of \var{bytes}; so shall $\var{k}
  + 1$.}
   
These retrieve and set two-byte representations of numbers at indices
\var{k} and $\var{k}+1$, according to the endianness specified by
\var{endianness}. The procedures with {\cf u16} in their names deal with the
unsigned representation; those with {\cf s16} in their names deal
with the two's complement representation.

The procedures with {\cf native} in their names employ the native
endianness, and only work at aligned indices:
\var{k} shall be a multiple of 2.
   
The \ldots{\cf -set!} procedures return \unspecifiedreturn.

\begin{scheme}
(define b
  (u8-list->bytes
    '(255 255 255 255 255 255 255 255
      255 255 255 255 255 255 255 253)))

(bytes-u16-ref b 14 (endianness little)) \lev 65023
(bytes-s16-ref b 14 (endianness little)) \lev -513
(bytes-u16-ref b 14 (endianness big)) \lev 65533
(bytes-s16-ref b 14 (endianness big)) \lev -3

(bytes-u16-set! b 0 12345 (endianness little))
(bytes-u16-ref b 0 (endianness little)) \lev 12345

(bytes-u16-native-set! b 0 12345)
(bytes-u16-native-ref b 0) \ev 12345

(bytes-u16-ref b 0 (endianness little)) \lev \unspecified
\end{scheme}
\end{entry}

\begin{entry}{%
\proto{bytes-u32-ref}{ bytes k endianness}{procedure}
\proto{bytes-s32-ref}{ bytes k endianness}{procedure}
\proto{bytes-u32-native-ref}{ bytes k}{procedure}
\proto{bytes-s32-native-ref}{ bytes k}{procedure}
\proto{bytes-u32-set!}{ bytes k n endianness}{procedure}
\proto{bytes-s32-set!}{ bytes k n endianness}{procedure}
\proto{bytes-u32-native-set!}{ bytes k n}{procedure}
\proto{bytes-s32-native-set!}{ bytes k n}{procedure}}
   
\domain{$\{\var{k}, \ldots, \var{k}+ 3\}$ shall be valid indices of \var{bytes}.}.
   
These retrieve and set four-byte representations of numbers at indices $\{\var{k},
\ldots, \var{k}+ 3\}$, according to the endianness specified by \var{endianness}. The
procedures with {\cf u32} in their names deal with the unsigned representation,
those with {\cf s32} with the two's complement representation.
   
The procedures with {\cf native} in their names employ the native endianness, and
only work at aligned indices: \var{k} shall be a multiple of 4..
   
The \ldots{\cf{}-set!} procedures return \unspecifiedreturn.

\begin{scheme}
(define b
  (u8-list->bytes
    '(255 255 255 255 255 255 255 255
      255 255 255 255 255 255 255 253)))

(bytes-u32-ref b 12 (endianness little)) \lev 4261412863
(bytes-s32-ref b 12 (endianness little)) \lev -33554433
(bytes-u32-ref b 12 (endianness big)) \lev 4294967293
(bytes-s32-ref b 12 (endianness big)) \lev -3
\end{scheme}
\end{entry}

\begin{entry}{%
\proto{bytes-u64-ref}{ bytes k endianness}{procedure}
\proto{bytes-s64-ref}{ bytes k endianness}{procedure}
\proto{bytes-u64-native-ref}{ bytes k}{procedure}
\proto{bytes-s64-native-ref}{ bytes k}{procedure}
\proto{bytes-u64-set!}{ bytes k n endianness}{procedure}
\proto{bytes-s64-set!}{ bytes k n endianness}{procedure}
\proto{bytes-u64-native-set!}{ bytes k n}{procedure}
\proto{bytes-s64-native-set!}{ bytes k n}{procedure}}
 
\domain{$\{\var{k}, \ldots, \var{k}+ 7\}$ shall be valid indices of \var{bytes}.}
   
These retrieve and set eight-byte representations of numbers at
indices $\{\var{k}, \ldots, \var{k}+ 7\}$, according to the endianness
specified by \var{endianness}. The procedures with {\cf u64} in their names deal
with the unsigned representation, those with {\cf s64} with the two's
complement representation.
   
The procedures with {\cf native} in their names employ the native endianness, and
only work at aligned indices: \var{k} shall be a multiple of 8.
   
The \ldots{\cf{}-set!} procedures return \unspecifiedreturn.

\begin{scheme}
(define b
  (u8-list->bytes
    '(255 255 255 255 255 255 255 255
      255 255 255 255 255 255 255 253)))

(bytes-u64-ref b 8 (endianness little)) \lev 18302628885633695743
(bytes-s64-ref b 8 (endianness little)) \lev -144115188075855873
(bytes-u64-ref b 8 (endianness big)) \lev 18446744073709551613
(bytes-s64-ref b 8 (endianness big)) \lev -3
\end{scheme}
\end{entry}

\begin{entry}{%
\proto{bytes=?}{ \vari{bytes} \varii{bytes}}{procedure}}
   
Returns \schtrue{} if \vari{bytes} and \varii{bytes} are equal---that
is, if they have the same length and equal bytes at all valid indices.
It returns \schfalse{} otherwise.
\end{entry}

\begin{entry}{%
\proto{bytes-ieee-single-native-ref}{ bytes k}{procedure}
\proto{bytes-ieee-single-ref}{ bytes k endianness}{procedure}}

\domain{$\{\var{k}, \ldots, \var{k}+3\}$ shall be valid indices of
  \var{bytes}.  For {\cf bytes-ieee-single-native-ref}, \var{k} shall
  be a multiple of $4$.}

These procedures return the inexact real that best represents the IEEE-754 single
precision number represented by the four bytes beginning at index
\var{k}.
\end{entry}

\begin{entry}{%
\proto{bytes-ieee-double-native-ref}{ bytes k}{procedure}
\proto{bytes-ieee-double-ref}{ bytes k endianness}{procedure}}

\domain{$\{\var{k}, \ldots, \var{k}+7\}$ shall be valid indices of
  \var{bytes}.  For {\cf bytes-ieee-double-native-ref}, \var{k} shall
  be a multiple of $8$.}

These procedures return the inexact real that best represents the IEEE-754 single
precision number represented by the eight bytes beginning at index
\var{k}.
\end{entry}

\begin{entry}{%
\proto{bytes-ieee-single-native-set!}{ bytes k x}{procedure}
\proto{bytes-ieee-single-set!}{ bytes k x endianness}{procedure}}

\domain{$\{\var{k}, \ldots, \var{k}+3\}$ shall be valid indices of
  \var{bytes}.  For {\cf bytes-ieee-single-native-set!}, \var{k} shall
  be a multiple of $4$.  \var{X} shall be a real number.}

These procedures store an IEEE-754 single precision representation of \var{x} into
elements \var{k} through $\var{k}+3$ of \var{bytes}, and returns
\unspecifiedreturn.
\end{entry}

\begin{entry}{%
\proto{bytes-ieee-double-native-set!}{ bytes k x}{procedure}
\proto{bytes-ieee-double-set!}{ bytes k x endianness}{procedure}}

\domain{$\{\var{k}, \ldots, \var{k}+7\}$ shall be valid indices of
  \var{bytes}.  For {\cf bytes-ieee-double-native-set!}, \var{k} shall
  be a multiple of $8$.}

These procedures store an IEEE-754 double precision representation of \var{x} into
elements \var{k} through $\var{k}+7$ of \var{bytes}, and returns
\unspecifiedreturn.
\end{entry}

\begin{entry}{%
\proto{bytes-copy!}{ source source-start target target-start n}{procedure}}

\domain{\var{Source-start}, \var{target-start},
  and \var{n} shall be non-negative exact integers that satisfy
  
  \begin{displaymath}
    \begin{array}{rcccccl}
      0 & \leq & \var{source-start} & \leq & \var{source-start} + \var{n} & \leq & l_{\var{source}}
      \\
      0 & \leq & \var{target-start} & \leq & \var{target-start} + \var{n} & \leq & l_{\var{target}}
    \end{array}
  \end{displaymath}
  %
  where $l_{\var{source}}$ is the length of \var{source} and
  $l_{\var{target}}$ is the length of \var{target}.}
   
   
  The {\cf bytes-copy!} procedure copies the bytes from \var{source} at indices 
  \begin{displaymath}
     \{ \var{source-start}, \ldots \var{source-start} + \var{n} - 1 \}
  \end{displaymath}
  to consecutive indices in \var{target} starting at \var{target-index}.
   
  This must work even if the memory regions for the source and the target
  overlap, i.e., the bytes at the target location after the copy must be
  equal to the bytes at the source location before the copy.
   
  This returns \unspecifiedreturn.
\begin{scheme}
(let ((b (u8-list->bytes '(1 2 3 4 5 6 7 8))))
  (bytes-copy! b 0 b 3 4)
  (bytes->u8-list b)) \ev (1 2 3 1 2 3 4 8)
\end{scheme}
\end{entry}

\begin{entry}{%
\proto{bytes-copy}{ bytes}{procedure}}
   
Returns a newly allocated copy of \var{bytes}.
\end{entry}

\begin{entry}{%
\proto{bytes->u8-list}{ bytes}{procedure}
\proto{u8-list->bytes}{ list}{procedure}}
   
\domain{\var{List} shall be a list of octets.}

The {\cf bytes->u8-list} procedure returns a newly allocated list of the bytes of
\var{bytes} in the same order.

The {\cf u8-list->bytes} procedure returns a newly allocated bytes object whose
elements are the elements of list \var{list}, in
the same order.  Analogous to {\cf list->vector}.
\end{entry}

\begin{entry}{%
\proto{bytes->uint-list}{ bytes endianness size}{procedure}
\proto{bytes->sint-list}{ bytes endianness size}{procedure}
\proto{uint-list->bytes}{ list endianness size}{procedure}
\proto{sint-list->bytes}{ list endianness size}{procedure}}
   
\domain{\var{Size} shall be a positive exact integer.}
   
These procedures convert between lists of integers and their consecutive
representations according to \var{size} and \var{endianness} in the
\var{bytes} objects in the same way as {\cf bytes->u8-list} and {\cf
  u8-list->bytes} do for one-byte representations.

\begin{scheme}
(let ((b (u8-list->bytes '(1 2 3 255 1 2 1 2))))
  (bytes->sint-list b (endianness little) 2)) \lev (513 -253 513 513)

(let ((b (u8-list->bytes '(1 2 3 255 1 2 1 2))))
  (bytes->uint-list b (endianness little) 2)) \lev (513 65283 513 513)
\end{scheme}
\end{entry}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "r6rs"
%%% End: 
