\documentclass[twoside,twocolumn]{algol60}
%\documentclass[twoside]{algol60}

\pagestyle{headings}
\showboxdepth=0
\makeindex
\input{commands}
\input{semantics-commands}

\texonly
\usepackage{xr}
\externaldocument[lib:]{r6rs-lib}
\endtexonly

\def\headertitle{Revised$^{5.93}$ Scheme}
\def\integerversion{6}

\begin{document}

\thispagestyle{empty}

\topnewpage[{
\begin{center}   {\huge\bf
        Revised{\Huge$^{\mathbf{\htmlonly\tiny\endhtmlonly{}5.93}}$} Report on the Algorithmic Language \\
                              \vskip 3pt
                                Scheme}

\vskip 1ex
$$
\begin{tabular}{l@{\extracolsep{.5in}}lll}
\multicolumn{4}{c}{M\authorsc{ICHAEL} S\authorsc{PERBER}}
\\
\multicolumn{4}{c}{W\authorsc{ILLIAM} C\authorsc{LINGER},
  R.\ K\authorsc{ENT} D\authorsc{YBVIG},
  M\authorsc{ATTHEW} F\authorsc{LATT},
  A\authorsc{NTON} \authorsc{VAN} S\authorsc{TRAATEN}}
\\
\multicolumn{4}{c}{(\textit{Editors})} \\
\multicolumn{4}{c}{
  R\authorsc{ICHARD} K\authorsc{ELSEY}, W\authorsc{ILLIAM} C\authorsc{LINGER},
  J\authorsc{ONATHAN} R\authorsc{EES}} \\
\multicolumn{4}{c}{(\textit{Editors, Revised$^5$ Report on the
    Algorithmic Language Scheme})} \\
\multicolumn{4}{c}{
  R\authorsc{OBERT} B\authorsc{RUCE} F\authorsc{INDLER}, J\authorsc{ACOB} M\authorsc{ATTHEWS}} \\
\multicolumn{4}{c}{(\textit{Authors, formal semantics})} \\[1ex]
\multicolumn{4}{c}{\bf 22 May 2007}
\end{tabular}
$$



\end{center}

\chapter*{Summary}
\medskip

The report gives a defining description of the programming language
Scheme.  Scheme is a statically scoped and properly tail-recursive
dialect of the Lisp programming language invented by Guy Lewis
Steele~Jr.\ and Gerald Jay~Sussman.  It was designed to have an
exceptionally clear and simple semantics and few different ways to
form expressions.  A wide variety of programming paradigms, including
imperative, functional, and message passing styles, find convenient
expression in Scheme.

The introduction offers a brief history of the language and of
the report.  It also contains a short tutorial to the basic concepts
of the language.

\begin{itemize}
\item Chapter~\ref{semanticchapter} gives an overview of the language.
\item Chapter~\ref{numbertypeschapter} explains Scheme's
  representation for numbers.
\item Chapter~\ref{readsyntaxchapter} defines the lexical syntax of
  Scheme and the syntax of Scheme forms.
\item Chapter~\ref{basicchapter} presents the fundamental semantic
  ideas of the language.
\item Chapter~\ref{terminologychapter} defines notational conventions
  used in the rest of the report.
\item Chapters~\ref{librarychapter} and \ref{programchapter} describe
  libraries and top-level programs, the basic organizational units of
  Scheme programs.
\item Chapter~\ref{expansionchapter} explains the macro-expansion process
  for Scheme code.
\item Chapter~\ref{baselibrarychapter} explains the Scheme base
  library which contains the fundamental forms useful to programmers.
\item Appendix~\ref{formalsemanticschapter} provides a formal
  semantics for a core of Scheme.
\item Appendix~\ref{derivedformsappendix} contains definitions for
  some of the derived forms described in the report.
\item Appendix~\ref{additionalmaterialappendix} describes some sources
  for additional material about Scheme.
\item Appendix~\ref{exampleappendix} contains an example of a small
  complete Scheme program.
\item Appendix~\ref{languagechangesappendix} describes changes to the
  language made since the previous revision of this report.
\end{itemize}

The report concludes with a list of references and an
alphabetic index.

This report is accompanied by a report describing standard
libraries~\cite{R6RS-libraries}; references to this document are
identified by designations such as ``library section'' or ``library
chapter''.  It is also accompanied by a report containing
non-normative appendices~\cite{R6RS-appendices}.

\bigskip

\begin{center}
{\large \bf
*** DRAFT*** \\
}\end{center}

This is a preliminary draft.  It is intended to reflect the decisions
taken by the editors' committee, but contains many mistakes,
ambiguities and inconsistencies.

}]

\texonly\clearpage\endtexonly

\chapter*{Contents}
\addvspace{3.5pt}                  % don't shrink this gap
\renewcommand{\tocshrink}{-4.0pt}  % value determined experimentally
{
\tableofcontents
}

\vfill
\eject


\input{intro}   \par
\vskip 2ex
\clearchaptergroupstar{Description of the language} %\unskip\vskip -2ex
\input{struct}  \par
\input{numbers} \par
\input{lex}     \par
\input{basic}   \par
\input{terminology} \par
\input{library} \par
\input{programs} \par
\input{expansion} \par
\input{base}    \par
\clearchaptergroupstar{Appendices}
\appendix
\chapter{Formal semantics}
\label{formalsemanticschapter}
\input{semantics} \par
\input{derived} \par
\input{repository} \par
\input{example} \par
\input{changes} \par
%\newpage                   %  Put bib on it's own page (it's just one)
%\twocolumn[\vspace{-.18in}]%  Last bib item was on a page by itself.
\renewcommand{\bibname}{References}

\bibliographystyle{plain}
\bibliography{abbrevs,rrs}

\vfill\eject


\newcommand{\indexheading}{Alphabetic index of definitions of
  concepts, keywords, and procedures}
\texonly
\newcommand{\indexintro}{The index includes entries from the library
  document; the entries are marked with ``(library)''.}
\endtexonly

\printindex

\end{document}
