\documentclass[twoside,twocolumn]{algol60}
%\documentclass[twoside]{algol60}

\usepackage{xr}

\pagestyle{headings} 
\showboxdepth=0
\makeindex
\input{commands}

\texonly\externaldocument[report:]{r6rs}\endtexonly

\def\headertitle{Revised$^{5.93}$ Scheme Libraries}
\def\integerversion{6}

\begin{document}

\thispagestyle{empty}

\topnewpage[{
\begin{center}   {\huge\bf
        Revised{\Huge$^{\mathbf{\htmlonly\tiny\endhtmlonly{}5.93}}$} Report on the Algorithmic Language \\
                              \vskip 3pt
                              Scheme\\
                                \vskip 1.5ex
                              --- Standard Libraries ---}

\vskip 1ex
$$
\begin{tabular}{l@{\extracolsep{.5in}}lll}
\multicolumn{4}{c}{M\authorsc{ICHAEL} S\authorsc{PERBER}}
\\
\multicolumn{4}{c}{W\authorsc{ILLIAM} C\authorsc{LINGER},
  R.\ K\authorsc{ENT} D\authorsc{YBVIG},
  M\authorsc{ATTHEW} F\authorsc{LATT},
  A\authorsc{NTON} \authorsc{VAN} S\authorsc{TRAATEN}}
\\
\multicolumn{4}{c}{(\textit{Editors})} \\
\multicolumn{4}{c}{
  R\authorsc{ICHARD} K\authorsc{ELSEY}, W\authorsc{ILLIAM} C\authorsc{LINGER},
  J\authorsc{ONATHAN} R\authorsc{EES}} \\
\multicolumn{4}{c}{(\textit{Editors, Revised$^5$ Report on the
    Algorithmic Language Scheme})} \\
\multicolumn{4}{c}{\bf 22 May 2007}
\end{tabular}
$$



\end{center}

\chapter*{Summary}
\medskip

The report gives a defining description of the standard libraries of the
programming language Scheme.

\begin{itemize}
\item 
Chapter~\ref{unicodechapter} describes the library implementing Unicode
semantics for characters and strings.
\item Chapter~\ref{bytevectorschapter}
describes the library for handling binary data.
\item 
Chapter~\ref{listutilities} describes the library containing list utilities
procedures.
\item Chapter~\ref{sortingchapter} describes the sorting
library.
\item
Chapter~\ref{controllib} describes the a library with various
control structures.
\item Chapter~\ref{recordschapter} describes the record system.
\item Chapter \ref{exceptionsconditionschapter} describes the libraries for
exceptions and conditions.
\item Chapter~\ref{iochapter} describes the I/O
libraries.
\item Chapter~\ref{filesystemchapter} describes the file-system
  library.
\item Chapter~\ref{programlibchapter} describes the library that allow
  programs to access command-lines and exit.
\item 
Chapter~\ref{numberchapter} describes specialized libraries
for dealing with numbers and arithmetic.
\item 
Chapter~\ref{syntaxcasechapter} describes the {\cf syntax-case}
facility for writing arbitrary macros.
\item Chapter~\ref{hashtablechapter}
describes the library for hashtables.
\item 
Chapter~\ref{enumerationschapter} describes the enumerations library.
\item Chapter~\ref{complibchapter} describes the composite library
containing most of the forms described in this report.
\item 
Chapter~\ref{evalchapter} describes the {\cf eval} facility for
evaluating Scheme expressions represented as data.
\item 
Chapter~\ref{pairmutationchapter} describes the operations for
mutating pairs.
\item 
Chapter~\ref{stringmutationchapter} describes the operation for
mutating strings.
\item Chapter~\ref{r5rscompatchapter} describes a library
with some procedures from the previous
version of this report for backwards compatibility.
\end{itemize}

The report concludes with a list of references and an
alphabetic index.

This report frequently refers back to the \textit{Revised$^6$ Report
  on the Algorithmic Language Scheme}~\cite{R6RS}; references to the
report are identified by designations such as ``report section'' or
``report chapter''.

\bigskip

\begin{center}
{\large \bf
*** DRAFT*** \\
}\end{center}

This is a preliminary draft.  It is intended to reflect the decisions
taken by the editors' committee, but contains many mistakes,
ambiguities and inconsistencies.

}]

\texonly\clearpage\endtexonly

\chapter*{Contents}
\addvspace{3.5pt}                  % don't shrink this gap
\renewcommand{\tocshrink}{-4.0pt}  % value determined experimentally
{%\footnotesize
\tableofcontents
}

\vfill

\texonly\clearpage\endtexonly

\input{unicode} \par
\input{bytevector} \par
\input{list}    \par
\input{sort}    \par
\input{control} \par
\input{records} \par
\input{exc}     \par
\input{io}      \par
\input{files}   \par
\input{programlib} \par
\input{arith}   \par
\input{syntax-case} \par
\input{hashtable} \par
\input{enum}    \par
\input{complib} \par
\input{eval}    \par
\input{setcar}  \par
\input{stringset} \par
\input{r5rscompat} \par
\input{libderived} \par
%\newpage                   %  Put bib on it's own page (it's just one)
%\twocolumn[\vspace{-.18in}]%  Last bib item was on a page by itself.
\renewcommand{\bibname}{References}

\bibliographystyle{plain}
\bibliography{abbrevs,rrs}

\vfill\eject


\newcommand{\indexheading}{Alphabetic index of definitions of
  concepts, keywords, and procedures}
\newcommand{\indexintro}{}

\printindex

\end{document}
